---
title: "Statistical Analysis (DESeq2, ALS vs Healthy Control)"
author: "Sushant Agellu"
date: "2026-04-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(data.table)
library(dplyr)
library(DESeq2)
library(tibble)

select  <- dplyr::select
filter  <- dplyr::filter
mutate  <- dplyr::mutate
arrange <- dplyr::arrange
```



# Paths — Update These to Match Your Machine

```{r paths}
clean_dir    <- "D:/sushant masters at SLU/Data analysis/001_Data_analysis/01_DataCleaning"

# Base output folder
output_dir <- "D:/sushant masters at SLU/Data analysis/001_Data_analysis/03_DataAnalysis/01_StatisticalAnalysis"

# Subfolders (all numbered for consistency)
results_dir     <- file.path(output_dir, "01_DESeq2_results")
vst_dir         <- file.path(output_dir, "02_VST_counts")
dispersion_dir  <- file.path(output_dir, "03_Dispersion_plots")

for (d in c(output_dir, results_dir, vst_dir, dispersion_dir)) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
}


```



# Load Cleaned Data

```{r load-data}
counts_clean <- fread(file.path(clean_dir, "GSE60424_counts_clean.csv"))
sample_map   <- fread(file.path(clean_dir, "GSE60424_sample_map.csv"))

counts_mat           <- as.data.frame(counts_clean)
rownames(counts_mat) <- counts_mat$gene_id
counts_mat$gene_id   <- NULL

cat("Counts matrix:", nrow(counts_mat), "genes ×", ncol(counts_mat), "samples\n")
cat("\nSamples per disease group:\n"); print(table(sample_map$diseasestatus_ch1))
cat("\nSamples per cell type:\n");     print(table(sample_map$celltype_ch1))
```


# Parameters

```{r parameters}
DISEASE_A   <- "ALS"
DISEASE_B   <- "Healthy Control"

CELL_TYPES  <- c("Neutrophils", "Monocytes", "CD4", "CD8", "B-cells", "NK", "Whole Blood")

MIN_COUNTS  <- 10
MIN_SAMPLES <- 2
PADJ_CUTOFF <- 0.05
LFC_CUTOFF  <- 1.0

safe <- function(x) gsub("[^A-Za-z0-9]", "_", x)
```



# Gene Symbol Annotation (Converts Ensembl IDs → gene symbols). Requires `org.Hs.eg.db` 

```{r annotation}
annotate_genes <- function(gene_ids) {
  gene_ids_clean <- sub("\\..*$", "", gene_ids)
  
  if (requireNamespace("org.Hs.eg.db", quietly = TRUE) &&
      requireNamespace("AnnotationDbi", quietly = TRUE)) {
    syms <- AnnotationDbi::mapIds(
      org.Hs.eg.db::org.Hs.eg.db,
      keys      = gene_ids_clean,
      column    = "SYMBOL",
      keytype   = "ENSEMBL",
      multiVals = "first"
    )
    result <- data.frame(gene_id     = gene_ids,
                         gene_symbol = unname(syms),
                         stringsAsFactors = FALSE)
    result$gene_symbol[is.na(result$gene_symbol)] <-
      result$gene_id[is.na(result$gene_symbol)]
    return(result)
  }
  
  data.frame(gene_id = gene_ids, gene_symbol = gene_ids, stringsAsFactors = FALSE)
}
```



# DESeq2 Analysis Function

#For each cell type:
#1. Subsets samples (ALS + HC only)
#2. Filters low-count genes (≥10 counts in ≥2 samples)
#3. Fits DESeq2 negative-binomial GLM with design = ~ disease
#4. Runs the Wald test, applies BH correction (padj)
#5. Saves: DEG table, VST counts, dispersion plot (PNG)

```{r deseq2-function}
run_deseq2 <- function(cell_type) {
  
  cat("\n", strrep("=", 60), "\n")
  cat("Cell type:", cell_type, "  |  ", DISEASE_A, "vs", DISEASE_B, "\n")
  cat(strrep("=", 60), "\n")
  
  meta_sub <- sample_map %>%
    filter(celltype_ch1      == cell_type,
           diseasestatus_ch1 %in% c(DISEASE_A, DISEASE_B)) %>%
    mutate(disease = factor(diseasestatus_ch1, levels = c(DISEASE_B, DISEASE_A)))
  
  if (nrow(meta_sub) < 4) { cat("  <4 samples — skipping.\n"); return(NULL) }
  
  shared_libs <- intersect(meta_sub$lib_id, colnames(counts_mat))
  counts_sub  <- counts_mat[, shared_libs, drop = FALSE]
  meta_sub    <- meta_sub %>% filter(lib_id %in% shared_libs) %>%
    arrange(match(lib_id, colnames(counts_sub)))
  stopifnot(all(colnames(counts_sub) == meta_sub$lib_id))
  
  cat("  Samples:", nrow(meta_sub),
      "( ALS =", sum(meta_sub$disease == DISEASE_A),
      "| HC =",  sum(meta_sub$disease == DISEASE_B), ")\n")
  
  keep        <- rowSums(counts_sub >= MIN_COUNTS) >= MIN_SAMPLES
  counts_filt <- counts_sub[keep, ]
  cat("  Genes after filtering:", nrow(counts_filt), "\n")
  
  col_data <- data.frame(disease = meta_sub$disease, row.names = meta_sub$lib_id)
  dds      <- DESeqDataSetFromMatrix(counts_filt, col_data, ~ disease)
  dds      <- DESeq(dds, quiet = TRUE)
  
  # Dispersion plot (PNG, QC)
  disp_file <- file.path(dispersion_dir, paste0("Dispersion_", safe(cell_type), ".png"))
  png(disp_file, width = 7, height = 5, units = "in", res = 300)
  plotDispEsts(dds, main = paste("Dispersion Estimates —", cell_type))
  dev.off()
  
  res    <- results(dds, contrast = c("disease", DISEASE_A, DISEASE_B), alpha = PADJ_CUTOFF)
  res_df <- as.data.frame(res) %>%
    rownames_to_column("gene_id") %>%
    filter(!is.na(padj)) %>%
    mutate(Expression = case_when(
      padj < PADJ_CUTOFF & log2FoldChange >  LFC_CUTOFF ~ "Up",
      padj < PADJ_CUTOFF & log2FoldChange < -LFC_CUTOFF ~ "Down",
      TRUE                                               ~ "NS"
    )) %>%
    arrange(padj)
  
  ann    <- annotate_genes(res_df$gene_id)
  res_df <- left_join(res_df, ann, by = "gene_id") %>%
    select(gene_id, gene_symbol, everything())
  
  cat("  DEGs (padj <", PADJ_CUTOFF, "& |LFC| >", LFC_CUTOFF, "):  Up =",
      sum(res_df$Expression == "Up"), "| Down =", sum(res_df$Expression == "Down"), "\n")
  
  ct_safe <- safe(cell_type)
  fwrite(res_df, file.path(results_dir, paste0("DESeq2_results_", ct_safe, ".csv")))
  
  vst_mat <- assay(vst(dds, blind = FALSE))
  vst_df  <- as.data.frame(vst_mat) %>% rownames_to_column("gene_id")
  fwrite(vst_df, file.path(vst_dir, paste0("VST_counts_", ct_safe, ".csv")))
  
  cat("  ✓ Saved: results, VST counts, dispersion plot\n")
  
  list(dds = dds, results = res_df, vst = vst_mat, meta = meta_sub)
}
```



# for running All 7 Cell Types

```{r run-all}
all_results <- list()
for (ct in CELL_TYPES) {
  all_results[[ct]] <- run_deseq2(ct)
}
cat("\n analysis complete.\n")
```



# Summary Table

```{r summary}
summary_df <- bind_rows(lapply(names(all_results), function(nm) {
  r <- all_results[[nm]]; if (is.null(r)) return(NULL)
  data.frame(
    CellType   = nm,
    N_Samples  = nrow(r$meta),
    N_Genes    = nrow(r$results),
    N_Up       = sum(r$results$Expression == "Up"),
    N_Down     = sum(r$results$Expression == "Down"),
    N_Total    = sum(r$results$Expression %in% c("Up", "Down")),
    Top_Gene   = r$results$gene_symbol[1]
  )
})) %>% arrange(desc(N_Total))

fwrite(summary_df, file.path(output_dir, "DEG_summary.csv"))
knitr::kable(summary_df, caption = "DESeq2 DEG Summary — ALS vs Healthy Control (sorted by total DEGs)")
```

