---
title: "02 — Visualization (ALS vs Healthy Control)"
author: "Sushant Agellu"
date: "2026-04-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 9, fig.height = 6)

library(data.table)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(RColorBrewer)
library(DESeq2)
library(tibble)
library(grid)

select  <- dplyr::select
filter  <- dplyr::filter
mutate  <- dplyr::mutate
arrange <- dplyr::arrange
```


# Paths — Update These to Match Your Machine

```{r paths}
clean_dir   <- "D:/sushant masters at SLU/Data analysis/001_Data_analysis/01_DataCleaning"

# Input: outputs from 01_StatisticalAnalysis 
stats_dir   <- "D:/sushant masters at SLU/Data analysis/001_Data_analysis/03_DataAnalysis/01_StatisticalAnalysis"
results_dir <- file.path(stats_dir, "01_DESeq2_results")
vst_dir     <- file.path(stats_dir, "02_VST_counts")

# Output folder 
plot_dir    <- "D:/sushant masters at SLU/Data analysis/001_Data_analysis/03_DataAnalysis/02_Visualization"
if (!dir.exists(plot_dir)) dir.create(plot_dir, recursive = TRUE)
```


# Parameters

```{r parameters}
DISEASE_A    <- "ALS"
DISEASE_B    <- "Healthy Control"

COLOR_ALS    <- "#984EA3"
COLOR_HC     <- "#377EB8"
COLOR_UP     <- "#E41A1C"
COLOR_DOWN   <- "#377EB8"
COLOR_NS     <- "grey70"

PADJ_CUTOFF  <- 0.05
LFC_CUTOFF   <- 1.0
N_LABEL      <- 15

CELL_TYPES   <- c("Neutrophils", "Monocytes", "CD4", "CD8", "B-cells", "NK", "Whole Blood")
FOCUS_CELLS  <- c("Neutrophils", "Monocytes")   # top 2 by DEG count

# PNG export settings
PNG_DPI      <- 300

safe <- function(x) gsub("[^A-Za-z0-9]", "_", x)

load_results <- function(cell_type) {
  f <- file.path(results_dir, paste0("DESeq2_results_", safe(cell_type), ".csv"))
  if (!file.exists(f)) { message("Missing: ", f); return(NULL) }
  fread(f)
}
load_vst <- function(cell_type) {
  f <- file.path(vst_dir, paste0("VST_counts_", safe(cell_type), ".csv"))
  if (!file.exists(f)) { message("Missing: ", f); return(NULL) }
  df <- as.data.frame(fread(f))
  rownames(df) <- df$gene_id
  df$gene_id <- NULL
  df
}

# Convenience wrapper for PNG saves
save_png <- function(plot, filename, width, height) {
  ggsave(file.path(plot_dir, filename), plot,
         width = width, height = height, dpi = PNG_DPI, bg = "white")
}
```



# Loading the Shared Data

```{r load-shared}
counts_clean <- fread(file.path(clean_dir, "GSE60424_counts_clean.csv"))
sample_map   <- fread(file.path(clean_dir, "GSE60424_sample_map.csv"))

counts_mat           <- as.data.frame(counts_clean)
rownames(counts_mat) <- counts_mat$gene_id
counts_mat$gene_id   <- NULL
```



#Sample Counts

```{r plot01-sample-counts, fig.height=5}
meta_sub <- sample_map %>%
  filter(diseasestatus_ch1 %in% c(DISEASE_A, DISEASE_B))

p01 <- ggplot(meta_sub, aes(x = celltype_ch1, fill = diseasestatus_ch1)) +
  geom_bar(position = position_dodge(width = 0.8), width = 0.7, colour = "white") +
  geom_text(stat = "count", aes(label = after_stat(count)),
            position = position_dodge(width = 0.8),
            vjust = -0.3, size = 3.8, fontface = "bold") +
  scale_fill_manual(values = setNames(c(COLOR_ALS, COLOR_HC), c(DISEASE_A, DISEASE_B))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Sample Counts: ALS vs Healthy Control",
       x = "Cell Type", y = "Number of Samples", fill = NULL) +
  theme_classic(base_size = 13) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "top")

print(p01)
save_png(p01, "01_SampleCounts.png", 8, 5)
```



# for the  Global PCA

```{r plot02-global-pca}
meta_all <- sample_map %>% filter(diseasestatus_ch1 %in% c(DISEASE_A, DISEASE_B))
libs_all   <- intersect(meta_all$lib_id, colnames(counts_mat))
counts_all <- counts_mat[, libs_all, drop = FALSE]
meta_all   <- meta_all %>% filter(lib_id %in% libs_all) %>%
  arrange(match(lib_id, colnames(counts_all)))

keep_all <- rowSums(counts_all >= 10) >= 3
dds_all  <- DESeqDataSetFromMatrix(
  counts_all[keep_all, ],
  colData = data.frame(disease  = factor(meta_all$diseasestatus_ch1),
                       celltype = factor(meta_all$celltype_ch1),
                       row.names = meta_all$lib_id),
  design  = ~ 1
)
vst_all  <- assay(vst(dds_all, blind = TRUE))

top_var  <- order(rowVars(vst_all), decreasing = TRUE)[1:2000]
pca_res  <- prcomp(t(vst_all[top_var, ]), scale. = FALSE)
pct_var  <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 1)
pca_df   <- data.frame(pca_res$x[, 1:2],
                       Disease  = meta_all$diseasestatus_ch1,
                       CellType = meta_all$celltype_ch1)

p02 <- ggplot(pca_df, aes(PC1, PC2, colour = CellType, shape = Disease)) +
  geom_point(size = 3.8, alpha = 0.9) +
  scale_colour_brewer(palette = "Set1") +
  scale_shape_manual(values = c(16, 17)) +
  labs(title    = "Global PCA — All Samples",
       subtitle = "Dominant variation is driven by cell type, not disease status",
       x = paste0("PC1 (", pct_var[1], "%)"),
       y = paste0("PC2 (", pct_var[2], "%)")) +
  theme_classic(base_size = 13)

print(p02)
save_png(p02, "02_PCA_Global.png", 8, 6)
```



# Showing of the DEG Summary Bar

```{r plot03-deg-summary, fig.height=5.5}
summary_rows <- list()
for (ct in CELL_TYPES) {
  r <- load_results(ct); if (is.null(r)) next
  n_up   <- sum(r$Expression == "Up")
  n_down <- sum(r$Expression == "Down")
  summary_rows[[ct]] <- data.frame(
    CellType  = ct,
    Direction = c("Up", "Down"),
    Count     = c(n_up, -n_down),
    Label     = c(n_up, n_down)
  )
}
summary_df <- bind_rows(summary_rows)
order_ct <- summary_df %>% group_by(CellType) %>%
  summarise(tot = sum(abs(Count))) %>% arrange(desc(tot)) %>% pull(CellType)
summary_df$CellType <- factor(summary_df$CellType, levels = order_ct)

p03 <- ggplot(summary_df, aes(CellType, Count, fill = Direction)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = Label,
                vjust = ifelse(Count >= 0, -0.4, 1.3)),
            size = 4, fontface = "bold") +
  geom_hline(yintercept = 0, colour = "black", linewidth = 0.4) +
  scale_fill_manual(values = c(Up = COLOR_UP, Down = COLOR_DOWN)) +
  scale_y_continuous(expand = expansion(mult = c(0.15, 0.15)), labels = abs) +
  labs(title    = "DEG Counts Across Cell Types — ALS vs Healthy Control",
       subtitle = paste("padj <", PADJ_CUTOFF, "& |LFC| >", LFC_CUTOFF,
                        " — sorted by total DEGs"),
       x = "Cell Type", y = "Number of DEGs (Down ← → Up)", fill = NULL) +
  theme_classic(base_size = 13) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1),
        legend.position = "top")

print(p03)
save_png(p03, "03_DEG_Summary.png", 8.5, 5.5)
```



# Focused PCA of Neutrophils & Monocytes

```{r plot04-05-focus-pca, fig.height=5}
plot_focus_pca <- function(cell_type) {
  vst_focus  <- load_vst(cell_type)
  meta_focus <- sample_map %>%
    filter(lib_id %in% colnames(vst_focus),
           diseasestatus_ch1 %in% c(DISEASE_A, DISEASE_B)) %>%
    arrange(match(lib_id, colnames(vst_focus)))
  
  vst_mat  <- as.matrix(vst_focus[, meta_focus$lib_id])
  tv       <- order(rowVars(vst_mat), decreasing = TRUE)[1:2000]
  pca_f    <- prcomp(t(vst_mat[tv, ]), scale. = FALSE)
  pct_f    <- round(100 * pca_f$sdev^2 / sum(pca_f$sdev^2), 1)
  df_pca   <- data.frame(pca_f$x[, 1:2],
                         Disease = factor(meta_focus$diseasestatus_ch1,
                                          levels = c(DISEASE_B, DISEASE_A)))
  
  ggplot(df_pca, aes(PC1, PC2, colour = Disease, fill = Disease)) +
    stat_ellipse(geom = "polygon", alpha = 0.12, level = 0.85, linewidth = 0.8) +
    geom_point(size = 4.5, alpha = 0.9) +
    scale_colour_manual(values = setNames(c(COLOR_ALS, COLOR_HC), c(DISEASE_A, DISEASE_B))) +
    scale_fill_manual(  values = setNames(c(COLOR_ALS, COLOR_HC), c(DISEASE_A, DISEASE_B))) +
    labs(title = paste("PCA —", cell_type),
         subtitle = paste(DISEASE_A, "vs", DISEASE_B),
         x = paste0("PC1 (", pct_f[1], "%)"),
         y = paste0("PC2 (", pct_f[2], "%)")) +
    theme_classic(base_size = 13) +
    theme(legend.position = "top")
}

pca_numbers <- c("04", "05")
for (i in seq_along(FOCUS_CELLS)) {
  ct <- FOCUS_CELLS[i]
  p  <- plot_focus_pca(ct)
  print(p)
  save_png(p, paste0(pca_numbers[i], "_PCA_", safe(ct), ".png"), 6, 5)
}
```



#  Volcano Plots of Neutrophils & Monocytes

```{r plot06-07-volcano, fig.height=6.5}
plot_volcano <- function(cell_type) {
  r <- load_results(cell_type); if (is.null(r)) return(NULL)
  
  r <- r %>% mutate(
    neg_log10_padj = pmin(-log10(padj + 1e-300), 50),
    Expression     = factor(Expression, levels = c("Up", "Down", "NS")),
    label_name     = ifelse(is.na(gene_symbol) | gene_symbol == "",
                            gene_id, gene_symbol))
  
  top_up   <- r %>% filter(Expression == "Up")   %>% slice_head(n = 8)
  top_down <- r %>% filter(Expression == "Down") %>% slice_head(n = 8)
  top_lbl  <- bind_rows(top_up, top_down)
  
  sc  <- table(r$Expression)
  sub <- paste0("Up: ", sc["Up"], "  |  Down: ", sc["Down"],
                "  |  padj < ", PADJ_CUTOFF, " & |LFC| > ", LFC_CUTOFF)
  
  ggplot(r, aes(log2FoldChange, neg_log10_padj, colour = Expression)) +
    geom_point(alpha = 0.55, size = 1.5) +
    geom_vline(xintercept = c(-LFC_CUTOFF, LFC_CUTOFF), linetype = "dashed", colour = "grey40") +
    geom_hline(yintercept = -log10(PADJ_CUTOFF),         linetype = "dashed", colour = "grey40") +
    geom_text_repel(data = top_lbl, aes(label = label_name),
                    size = 3.2, max.overlaps = 30,
                    segment.colour = "grey50", fontface = "italic") +
    scale_colour_manual(values = c(Up = COLOR_UP, Down = COLOR_DOWN, NS = COLOR_NS)) +
    labs(title    = paste("Volcano —", cell_type),
         subtitle = sub,
         x = paste0("log2 Fold Change (", DISEASE_A, " / ", DISEASE_B, ")"),
         y = "-log10 (adjusted p-value)", colour = NULL) +
    theme_classic(base_size = 13) +
    theme(legend.position = "top")
}

vol_numbers <- c("06", "07")
for (i in seq_along(FOCUS_CELLS)) {
  ct <- FOCUS_CELLS[i]
  p  <- plot_volcano(ct)
  print(p)
  save_png(p, paste0(vol_numbers[i], "_Volcano_", safe(ct), ".png"), 7.5, 6.5)
}
```



#Top DEG Heatmaps (Neutrophils & Monocytes)

Each cell type gets its own heatmap of the top 30 DEGs (z-scored VST counts).

```{r plot08-09-heatmap, fig.height=9}
draw_top_heatmap <- function(cell_type, number_str, n_genes = 30) {
  r_tbl   <- load_results(cell_type)
  vst_mat <- as.matrix(load_vst(cell_type))
  if (is.null(r_tbl) || is.null(vst_mat)) return(NULL)
  
  top_genes <- r_tbl %>% filter(Expression != "NS") %>% slice_head(n = n_genes)
  if (nrow(top_genes) < 2) {
    message("Not enough DEGs for heatmap: ", cell_type); return(NULL)
  }
  
  meta_hm <- sample_map %>%
    filter(lib_id %in% colnames(vst_mat),
           diseasestatus_ch1 %in% c(DISEASE_A, DISEASE_B)) %>%
    arrange(diseasestatus_ch1)
  
  mat_hm <- vst_mat[top_genes$gene_id, meta_hm$lib_id, drop = FALSE]
  rn     <- top_genes$gene_symbol
  rn[is.na(rn) | rn == ""] <- top_genes$gene_id[is.na(rn) | rn == ""]
  rownames(mat_hm) <- make.unique(rn)
  
  mat_z <- t(scale(t(mat_hm)))
  mat_z <- pmax(pmin(mat_z, 3), -3)
  
  ann_col    <- data.frame(Disease = meta_hm$diseasestatus_ch1,
                           row.names = meta_hm$lib_id)
  ann_colors <- list(Disease = setNames(c(COLOR_ALS, COLOR_HC),
                                        c(DISEASE_A, DISEASE_B)))
  
  out_file <- file.path(plot_dir, paste0(number_str, "_Heatmap_", safe(cell_type), ".png"))
  
  # Save PNG
  pheatmap(
    mat_z,
    annotation_col    = ann_col,
    annotation_colors = ann_colors,
    color             = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
    cluster_rows      = TRUE,
    cluster_cols      = TRUE,
    show_colnames     = FALSE,
    fontsize_row      = 8,
    main              = paste("Top", n_genes, "DEGs —", cell_type, "— ALS vs HC"),
    filename          = out_file,
    width  = 7,
    height = 9
  )
  
  # Inline render
  grid.newpage()
  pheatmap(
    mat_z,
    annotation_col    = ann_col,
    annotation_colors = ann_colors,
    color             = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
    cluster_rows      = TRUE,
    cluster_cols      = TRUE,
    show_colnames     = FALSE,
    fontsize_row      = 8,
    main              = paste("Top", n_genes, "DEGs —", cell_type, "— ALS vs HC")
  )
}

draw_top_heatmap("Neutrophils", "08")
draw_top_heatmap("Monocytes",   "09")
```



# Consensus Heatmap (Neutrophils ∩ Monocytes shared DEGs)

#Reasoning part : Identifies genes significantly changed in BOTH Neutrophils AND Monocytes — shared signal is more likely to be a real ALS response (vs cell-type-specific noise). This is the strongest single evidence plot in the presentation.

```{r plot10-consensus-heatmap, fig.height=8}
# Get significant genes per focus cell type
deg_sets <- lapply(FOCUS_CELLS, function(ct) {
  r <- load_results(ct)
  r %>% filter(Expression != "NS") %>%
    mutate(CellType = ct) %>%
    select(gene_id, gene_symbol, log2FoldChange, padj, Expression, CellType)
})
names(deg_sets) <- FOCUS_CELLS

# Shared gene IDs (significant in BOTH)
shared_ids <- Reduce(intersect, lapply(deg_sets, function(d) d$gene_id))
cat("Shared DEGs between", paste(FOCUS_CELLS, collapse = " and "), ":",
    length(shared_ids), "\n")

if (length(shared_ids) >= 2) {
  
  # Build a matrix of log2 fold changes: rows = shared genes, cols = cell types
  lfc_mat <- sapply(FOCUS_CELLS, function(ct) {
    d <- deg_sets[[ct]]
    d$log2FoldChange[match(shared_ids, d$gene_id)]
  })
  rownames(lfc_mat) <- shared_ids
  
  # Replace rownames with gene symbols
  sym_lookup <- deg_sets[[1]][match(shared_ids, deg_sets[[1]]$gene_id), ]
  rn <- sym_lookup$gene_symbol
  rn[is.na(rn) | rn == ""] <- shared_ids[is.na(rn) | rn == ""]
  rownames(lfc_mat) <- make.unique(rn)
  
  # Cap extreme values for display
  lfc_mat_disp <- pmax(pmin(lfc_mat, 6), -6)
  
  # Color breaks centered on zero
  max_abs <- max(abs(lfc_mat_disp), na.rm = TRUE)
  breaks  <- seq(-max_abs, max_abs, length.out = 101)
  
  out_file <- file.path(plot_dir, "10_Consensus_Heatmap_Neutrophils_Monocytes.png")
  
  pheatmap(
    lfc_mat_disp,
    color         = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
    breaks        = breaks,
    cluster_rows  = TRUE,
    cluster_cols  = FALSE,
    display_numbers = TRUE,
    number_format = "%.1f",
    number_color  = "black",
    fontsize_row  = 9,
    fontsize_col  = 12,
    main          = paste0("Consensus DEGs: Neutrophils ∩ Monocytes (n = ",
                           length(shared_ids), ") — log2 Fold Change"),
    filename      = out_file,
    width  = 6,
    height = max(4, length(shared_ids) * 0.28)
  )
  
  grid.newpage()
  pheatmap(
    lfc_mat_disp,
    color         = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
    breaks        = breaks,
    cluster_rows  = TRUE,
    cluster_cols  = FALSE,
    display_numbers = TRUE,
    number_format = "%.1f",
    number_color  = "black",
    fontsize_row  = 9,
    fontsize_col  = 12,
    main          = paste0("Consensus DEGs: Neutrophils ∩ Monocytes (n = ",
                           length(shared_ids), ") — log2 Fold Change")
  )
  
  # Also export the table
  consensus_df <- data.frame(
    gene_id      = shared_ids,
    gene_symbol  = rownames(lfc_mat),
    LFC_Neutrophils = lfc_mat[, "Neutrophils"],
    LFC_Monocytes   = lfc_mat[, "Monocytes"],
    stringsAsFactors = FALSE
  )
  fwrite(consensus_df, file.path(plot_dir, "10_Consensus_DEGs_Neutrophils_Monocytes.csv"))
  
} else {
  cat("Not enough shared DEGs to draw the consensus heatmap.\n")
}
```


# Appendix: Volcano Plots (remaining cell types)

```{r plot11-appendix-volcanos, fig.height=5.5}
for (ct in setdiff(CELL_TYPES, FOCUS_CELLS)) {
  r <- load_results(ct); if (is.null(r)) next
  
  r <- r %>% mutate(
    neg_log10_padj = pmin(-log10(padj + 1e-300), 50),
    Expression = factor(Expression, levels = c("Up", "Down", "NS")),
    label_name = ifelse(is.na(gene_symbol) | gene_symbol == "",
                        gene_id, gene_symbol))
  top_lbl <- r %>% filter(Expression != "NS") %>% slice_head(n = N_LABEL)
  sc  <- table(r$Expression)
  sub <- paste0("Up: ", sc["Up"], "  |  Down: ", sc["Down"])
  
  p <- ggplot(r, aes(log2FoldChange, neg_log10_padj, colour = Expression)) +
    geom_point(alpha = 0.5, size = 1.2) +
    geom_vline(xintercept = c(-LFC_CUTOFF, LFC_CUTOFF), linetype = "dashed", colour = "grey40") +
    geom_hline(yintercept = -log10(PADJ_CUTOFF),         linetype = "dashed", colour = "grey40") +
    geom_text_repel(data = top_lbl, aes(label = label_name),
                    size = 2.8, max.overlaps = 20, fontface = "italic") +
    scale_colour_manual(values = c(Up = COLOR_UP, Down = COLOR_DOWN, NS = COLOR_NS)) +
    labs(title = paste("Volcano —", ct), subtitle = sub,
         x = "log2 Fold Change", y = "-log10 (padj)", colour = NULL) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top")
  
  print(p)
  save_png(p, paste0("11_Volcano_", safe(ct), ".png"), 6.5, 5.5)
}
```

---

#Appendix: MA Plots (all cell types)

```{r plot12-appendix-ma, fig.height=5.5}
for (ct in CELL_TYPES) {
  r <- load_results(ct); if (is.null(r)) next
  
  r <- r %>% mutate(
    Expression = factor(Expression, levels = c("Up", "Down", "NS")),
    label_name = ifelse(is.na(gene_symbol) | gene_symbol == "",
                        gene_id, gene_symbol))
  top_lbl <- r %>% filter(Expression != "NS") %>% slice_head(n = 10)
  
  p <- ggplot(r, aes(log10(baseMean + 1), log2FoldChange, colour = Expression)) +
    geom_point(alpha = 0.4, size = 1.1) +
    geom_hline(yintercept = c(-LFC_CUTOFF, 0, LFC_CUTOFF),
               linetype   = c("dashed", "solid", "dashed"),
               colour     = c("grey40", "black", "grey40")) +
    geom_text_repel(data = top_lbl, aes(label = label_name),
                    size = 2.8, max.overlaps = 15, fontface = "italic") +
    scale_colour_manual(values = c(Up = COLOR_UP, Down = COLOR_DOWN, NS = COLOR_NS)) +
    labs(title = paste("MA Plot —", ct),
         x = "log10 (Mean Count + 1)", y = "log2 Fold Change", colour = NULL) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top")
  
  print(p)
  save_png(p, paste0("12_MA_", safe(ct), ".png"), 6.5, 5.5)
}
```


