---
title: "01_DataCleaning"
author: "Sushant Agellu"
date: "2026-02-25"
output: html_document
---

# Loading packages data.table for fast I/O, dplyr for manipulation, stringr for text cleaning
```{r setup, include=FALSE}

library(data.table)
library(dplyr)
library(stringr)

```



# Loading raw metadata and expression matrix downloaded from GEO website of the (GSE60424) data 

```{r}
gse_id <- "GSE60424"

meta_path  <- "D:/sushant masters at SLU/Data analysis/001_Data_analysis/00_DataFiles/GSE60424_metadata.csv"

expr_path <- "D:/sushant masters at SLU/Data analysis/001_Data_analysis/00_DataFiles/GSE60424_expression_matrix_from_supp.csv"

metadata_raw <- fread(meta_path)
expr_raw     <- fread(expr_path)

dim(metadata_raw)
dim(expr_raw)
```



# this would help clean the data that was downloaded ready for my next steps as there are lots of issues in raw data 
```{r}
# Make column names easier to work with
metadata <- metadata_raw
names(metadata) <- names(metadata) |>
  tolower() |>
  str_replace_all("[^a-z0-9]+", "_") |>
  str_replace_all("_+$", "")

# Common useful columns in this dataset
# (Some GEO exports store them as: age_ch1, diseasestatus_ch1, celltype_ch1, gender_ch1, etc.)

keep_cols <- c(
  "geo_accession", "title",
  "age_ch1", "gender_ch1", "race_ch1",
  "celltype_ch1", "diseasestatus_ch1",
  "donorid_ch1", "samplename_ch1",
  "collectiondate_ch1"
)

# Keeps only demographic, clinical, and sample-identifier columns needed downstream

keep_cols <- keep_cols[keep_cols %in% names(metadata)]
metadata_clean <- metadata[, ..keep_cols]

# Basic standardization (small, safe cleaning)
if ("gender_ch1" %in% names(metadata_clean)) {
  metadata_clean$gender_ch1 <- toupper(metadata_clean$gender_ch1)
}

if ("diseasestatus_ch1" %in% names(metadata_clean)) {
  metadata_clean$diseasestatus_ch1 <- str_squish(metadata_clean$diseasestatus_ch1)
}

# Save cleaned metadata
if (!dir.exists("01_DataCleaning")) dir.create("01_DataCleaning")
fwrite(metadata_clean, file.path("D:/sushant masters at SLU/Data analysis/001_Data_analysis/01_DataCleaning", paste0(gse_id, "_metadata_clean.csv")))
dim(metadata_clean)
head(metadata_clean, 5)
```





# Load expression matrix (raw) from 00_DataFiles

```{r}
gse_id <- "GSE60424"

counts <- fread(
  file.path("..", "00_DataFiles",
            paste0(gse_id, "_expression_matrix_from_supp.csv"))
)
```



##  Clean expression matrix (simple version)
```{r}


# Rename first column to gene_id (whatever its original name was)
names(counts)[1] <- "gene_id"

# Remove duplicate genes if any
counts_clean <- counts %>%
  distinct(gene_id, .keep_all = TRUE)

# Check structure
dim(counts_clean)

# Preview first few columns
head(counts_clean[, 1:6])

# Save cleaned counts
fwrite(counts_clean,
       file.path("D:/sushant masters at SLU/Data analysis/001_Data_analysis/01_DataCleaning", paste0(gse_id, "_counts_clean.csv")))
```

# To Build GSMlib_id mapping table — expression columns use lib IDs, metadata uses GSM IDs
```{r}
# Try extracting lib### from title (best case)
sample_map <- metadata_clean %>%
  mutate(
    lib_id = if ("title" %in% names(metadata_clean)) str_extract(title, "lib\\d+") else NA_character_
  ) %>%
  mutate(
    gsm_id = if ("geo_accession" %in% names(metadata_clean)) geo_accession else NA_character_
  ) %>%
  select(gsm_id, lib_id, everything())

found_lib <- sum(!is.na(sample_map$lib_id))

if (found_lib == 0) {
  # Fallback: assume lib columns are in the same order as metadata rows
  lib_order <- sample_cols
  if (nrow(metadata_clean) != length(lib_order)) {
    warning("Metadata rows and lib columns differ; order-based mapping may be unsafe.")
  }

  sample_map <- metadata_clean %>%
    mutate(
      gsm_id = if ("geo_accession" %in% names(metadata_clean)) geo_accession else as.character(row_number()),
      lib_id = lib_order[1:n()]
    ) %>%
    select(gsm_id, lib_id, everything())
}

# Save sample map
fwrite(sample_map, file.path("D:/sushant masters at SLU/Data analysis/001_Data_analysis/01_DataCleaning", paste0(gse_id, "_sample_map.csv")))

head(sample_map, 10)
cat("Lib IDs found from title:", found_lib, "\n")
```

